---
title: "Analysis of the Adverse Health and Economic Impacts of US Storms"
author: "Kieso Jan"
date: "7/23/2021"
output: html_document
---

## 1: Synopsis 

The goal of this analysis is to explore the NOAA Storm Database and explore the effects of severe weather events on both population and economy. The events in the database start in the year 1950 and end in November 2011. 
this analysis is aiming to answer:

1. Across the United States, which types of events are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

Information on the Data: [Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

```{r Environment Setup}

library(data.table)
library(ggplot2)
library(dplyr,warn.conflicts = FALSE)

```


```{r Data Download}

DatafileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(DatafileUrl, destfile = paste0("data", '/repdata_data_StormData.csv.bz2'))

```
```{r Data Loading}

stormDF <- read.csv("data/repdata_data_StormData.csv.bz2")

```

### 2.2: Examining Column Names

```{r ColumnNames}

colnames(stormDF)

```

since it's not all of columns will be used, Let's only keep columns relevant to following analysis:

1. EVTYPE 2.FATALITIES 3.INJURIES 4.PROPDMG 5.PROPDMGEXP 6.CROPDMG 7.CROPDMGEXP


### 2.3: Data Subsetting

Basically, we remove the columns we don't need for clarity in this analysis, and keep rows is not (1) unknown (2) value > 0

```{r DataSubsetting, results="hide"}

# Finding columns to keep
stormDF<- select(stormDF, c("EVTYPE"
                         , "FATALITIES"
                         , "INJURIES"
                         , "PROPDMG"
                         , "PROPDMGEXP"
                         , "CROPDMG"
                         , "CROPDMGEXP")) %>%
## Only use data where fatalities or injuries occurred.  
             filter(EVTYPE != "?" & 
             (INJURIES > 0 | FATALITIES > 0 | PROPDMG > 0 | CROPDMG > 0))

str(stormDF)

```

### 2.4: Converting Exponent Columns into Actual Exponents instead of (-,+, H, K, etc)

Based on the Explanation of document. Alphabetical characters used to signify magnitude include “K” for thousands, “M” for millions, and “B” for billions. 

```{r}

stormDF$PROPDMGEXP <-  toupper(stormDF$PROPDMGEXP)
stormDF$CROPDMGEXP <-  toupper(stormDF$CROPDMGEXP)

table(stormDF$PROPDMGEXP)
table(stormDF$CROPDMGEXP)

```


```{r ConvertExponents, results="hide"}

# Map property damage alphanumeric exponents to numeric values.

stormDF <- stormDF %>% mutate(PROPDMGEXP 
                          = case_when(
                           PROPDMGEXP == "\"\"" ~ 10^0,
                           PROPDMGEXP ==  "-" ~ 10^0,
                           PROPDMGEXP ==  "+" ~ 10^0,
                           PROPDMGEXP ==  "0" ~ 10^0,
                           PROPDMGEXP ==  "1" ~ 10^1,
                           PROPDMGEXP ==  "2" ~ 10^2,
                           PROPDMGEXP ==  "3" ~ 10^3,
                           PROPDMGEXP ==  "4" ~ 10^4,
                           PROPDMGEXP ==  "5" ~ 10^5,
                           PROPDMGEXP ==  "6" ~ 10^6,
                           PROPDMGEXP ==  "7" ~ 10^7,
                           PROPDMGEXP ==  "8" ~ 10^8,
                           PROPDMGEXP ==  "9" ~ 10^9,
                           PROPDMGEXP ==  "H" ~ 10^2,
                           PROPDMGEXP ==  "K" ~ 10^3,
                           PROPDMGEXP ==  "M" ~ 10^6,
                           PROPDMGEXP ==  "B" ~ 10^9
                           ))

stormDF <- stormDF %>% mutate(PROPDMGEXP = if_else(is.na(PROPDMGEXP),10^0,PROPDMGEXP))


# Map crop damage alphanumeric exponents to numeric values

stormDF <- stormDF %>% mutate(CROPDMGEXP 
                          = case_when(
                           CROPDMGEXP ==  "?" ~ 10^0,
                           CROPDMGEXP ==  "0" ~ 10^0,
                           CROPDMGEXP ==  "K" ~ 10^3,
                           CROPDMGEXP ==  "M" ~ 10^6,
                           CROPDMGEXP ==  "B" ~ 10^9
                           ))

stormDF <- stormDF %>% mutate(CROPDMGEXP = if_else(is.na(CROPDMGEXP),10^0,CROPDMGEXP))

table(stormDF$PROPDMGEXP)
table(stormDF$CROPDMGEXP)

```

### 2.5: Making Economic Cost Columns

```{r EconomicCostColumns}


stormDF <- stormDF %>% mutate(propCost = PROPDMG * PROPDMGEXP,
                              cropCost = CROPDMG * CROPDMGEXP)


```

### 2.6: Making a summary result for total cost  of Property and Crop 

```{r TotalPropertyCropCost}


totalCostDF <- stormDF %>% group_by(EVTYPE) %>%
  summarize(propCost = sum(propCost), cropCost = sum(cropCost), 
            Total_Cost  = sum(propCost) + sum(cropCost)) %>%
            arrange(desc(Total_Cost))

totalCostDF <- head(totalCostDF, 10)
                
head(totalCostDF, 5)


```

### ### 2.7:Making a summary result for  Total Fatalities and Injuries

```{r TotalFatalitiesInjuriesCalc}


totalInjuriesDT <- stormDF %>% group_by(EVTYPE) %>%
  summarize(FATALITIES = sum(FATALITIES), 
            INJURIES = sum(INJURIES), 
            totals  = sum(FATALITIES) + sum(INJURIES)) %>%
  arrange(desc(FATALITIES))

totalInjuriesDT <- head(totalInjuriesDT, 10)

head(totalInjuriesDT, 5)

```

## 3: Findings & Conclustions


### 3.1: Top 10 Events that are Most Harmful to Population Health 

Melting data format so that it is easier to put in bar graph format

```{r HealthOutputs}

terrible_things <- melt(totalInjuriesDT, id.vars="EVTYPE", variable.name = "terrible_thing")

head(terrible_things, 5)

```
```{r Chart of health}

ggplot(terrible_things, aes(x=reorder(EVTYPE, value), y=value)) +
  theme_bw() +
  # Plot data as bar chart
  geom_bar(stat="identity", aes(fill=terrible_thing), position="dodge") +  
  # Format y-axis scale and set y-axis label
  ylab("Frequency Count") +  xlab("Event Type") +                          
  # Rotate x-axis tick labels 
  theme(axis.text.x = element_text(angle=45, hjust=1)) +                   
  ggtitle("Top 10 US Killer Events") + 
  # Set chart title and center it
  theme(plot.title = element_text(hjust = 0.5)) + 
  # apply chart theme
  scale_fill_brewer(palette="Pastel2")  +    
  # setup legend
  theme(legend.position = "top", text = element_text(size = 10)) +
  coord_flip() 

```


### 3.2: Events that have the Greatest Economic Consequences

Melting data format so that it is easier to put in bar graph format 

```{r EconConsequences}

econ_consequences <- melt(totalCostDF, id.vars="EVTYPE", variable.name = "Damage_Type")

head(econ_consequences, 5)

```

```{r Chart of Economic Consequences}

  ggplot(econ_consequences, aes(x=reorder(EVTYPE, value), y=value)) +
  theme_bw() +
  # Plot data as bar chart
  geom_bar(stat="identity", aes(fill=Damage_Type), position="dodge") +  
  # Format y-axis scale and set y-axis label
  ylab("Cost (USD)") +  xlab("Event Type") +                          
  # Rotate x-axis tick labels 
  theme(axis.text.x = element_text(angle=45, hjust=1)) +                   
  # Set chart title and center it
  ggtitle("Top 10 US Storm Events causing Economic Consequences") + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  scale_fill_brewer(palette="Pastel2")  +    
  theme(legend.position = "top", text = element_text(size = 10)) +
  coord_flip() 

```












